---
title: "WORKFLOW for data analyses of Fuquene record"
author: 'Author: Xavier Benito (xavier.benito.granell@gmail.com)'
date: 'Date: 11/05/2022'
output:
  html_notebook: default
  pdf_document: default
---

## Introduction
This workflow is written as a R Markdown document to show initial exploratory analyses of paleoclimatic and paleoenvironmental evolution of Lake Fuquene (Colombia, 2540m elevation, 5°27′N, 73°46′W). This document shows the most relevant code. See R files within the folder "scripts" for a more specific code. 

### Study area and datasets description
+ **Lake Fuquene**: Pleistocene in origin, morrained-dammed tropical lake located in the Eastern Cordiellera of Colombian Andes. Water depth=2-6 m

### Three sediment cores, composite of 250 kyr diatom, geochemical and pollen record
+ **F9**: 60-m long, sampled for diatom, pollen and geochemistry
+ **F7**: sampled for diatoms
+ **M1**: sampled for diatoms and pollen


## Main numerical steps:
+ Clean data and estimate Principal Curves for each assemblage (pollen and diatoms)
+ Plot stratigraphical floristic changes over time to identify periods of change
+ Interpolate PrC from coarser to lower resolution 
+ GAM covariates model
+ HGAM-derivatives: assemblage-wide rate of change


### Load packages 
```{r packages, echo=TRUE, message=FALSE, warning=FALSE}
library(analogue)
library(tidyverse)
library(vegan)
library(analogue)
library(mgcv)
library(ggplot2)
library(gratia)
```

## Step 1: Clean data and estimate Principal Curves (PrC) of the diatom cores (M1, F7 and F9)
PrC is a nonlinear ordination technique that extracts a single gradient of variation from multivariate data.

```{r principal curves diatoms, echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
F_diatoms <- read.csv("data/Fuquene_diatoms_counts.csv") %>%
  rename(core=ï..core) #rename extraneous column name 

agedepth <- F_diatoms[, names(F_diatoms) %in% c("depth", "upper_age", "lower_age", "core")] 
diat <- F_diatoms[, !names(F_diatoms) %in% c("depth", "upper_age", "lower_age", "core")]
diat[is.na(diat)] <- 0

diatoms_save <- cbind(agedepth, diat)

#this is to transform to tidy format, calculate % and subset more common species
new <- diatoms_save %>% 
  gather(key = taxa, value = count, -depth, -upper_age, -lower_age, -core) %>%
  #mutate(taxa = plyr::mapvalues(taxa, from = changes[,1], to = changes$new_2)) %>%
  group_by(depth, taxa, upper_age, lower_age) %>%
  summarise(count = sum(count)) %>%
  filter(!count == "0" ) %>% #this is to remove empty samples (rows)
  filter(!upper_age==0.0) %>% #this is to drop extraneous ages
  ungroup() %>%
  group_by(depth) %>%
  mutate(relative_abundance_percent = count / sum(count) * 100) %>%
  ungroup()

# filter more abundant taxa; format need to be on long-wide format-->no spreaded 
core_common_taxa <- new %>%
  group_by(taxa) %>%
  summarise(max_rel_abund = max(relative_abundance_percent)) %>%
  filter(max_rel_abund >= 5) %>%
  arrange(max_rel_abund) %>%
  pull(taxa)

# select from initial table
core_counts_common <- new %>%
  filter(taxa %in% core_common_taxa) %>%
  mutate(taxa = factor(taxa, levels = core_common_taxa)) %>%
  arrange(taxa)

#make it wide
core_counts_wide_diatoms <- core_counts_common %>%
  select(depth, upper_age, lower_age, taxa, relative_abundance_percent) %>%
  spread(key = taxa, value = relative_abundance_percent) %>%
  arrange(upper_age) #sort by increasing time
```

### Check temporal resolution of diatom record
```{r}
median(diff(core_counts_wide_diatoms$upper_age))
plot.ts(diff(core_counts_wide_diatoms$upper_age))
```

### Estime PrC and plot trends
```{r PrC diatoms, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
## extract agedepth variables
agedepth <- core_counts_wide[,names(core_counts_wide) %in% c("depth", "upper_age", "lower_age")]
diatoms <- core_counts_wide[,!(names(core_counts_wide) %in% c("upper_age", "lower_age", "depth"))]

# Transform data to Hellinger form
diatoms[is.na(diatoms)] <- 0 #Replace NA (if any) by 0
diat_hell <- decostand(diatoms, method="hellinger")

# Run Principal Curves
diat_prc <- prcurve(diat_hell, method = "ca", trace = TRUE, vary = TRUE, penalty = 1.4)

## Extract position on the curve
scrs_prc <- scores(diat_prc, display = "curve")

# Combine dataframe with ages and depths
diatomsPrC <- cbind(agedepth, scrs_prc)

# Plot Pcurves with depth and ages
diat_plt_prc <- ggplot(diatomsPrC, aes(x = upper_age, y = PrC)) +
  geom_line() + geom_point() +
  labs(y = "PrC", x = "Age (cal yr BP)", title = "") +
  ggtitle("Diatoms PrC") +
  theme_bw()
diat_plt_prc

```

## Stratigraphic diatom plot

### First do CONISS classification to see statistically significant stratigraphic zones
```{r CONISS diatom, echo=FALSE, message=FALSE, warning=FALSE}
library(rioja)

#do coniss to add statistically significant stratigraphic zones
core_counts_wide[is.na(core_counts_wide)] <- 0
      
diatHel <- decostand(core_counts_wide[,4:ncol(core_counts_wide)], method="hellinger")
diss <- vegdist(diatHel, method="bray")
clust <- chclust(diss, method="coniss")
bstick(clust)

#doesn't work
zones <- cutree(clust, k=8)
locate <- cumsum(rle(zones)$lengths)+1
zones <- core_counts_wide_diatoms[locate, ][,2]
zones <- zones$upper_age

```

```{r stratiplot, results='markup'}
## Using tidypaleo R package (https://fishandwhistle.net/post/2018/stratigraphic-diagrams-with-tidypaleo-ggplot2/)
library(tidypaleo) #remotes::install_github("paleolimbot/tidypaleo")
library(patchwork)

theme_set(theme_bw(9))

# Plot  
diat_plot <- ggplot(core_counts_common, aes(x = relative_abundance_percent, y = upper_age)) +
  geom_col_segsh() +
  scale_y_reverse() +
  facet_abundanceh(vars(taxa)) +
  labs(x = "Relative abundance (%)", y = "Cal yr BP") +
  # add CONISS zones
  geom_hline(yintercept = zones, col = "blue", lty = 1, alpha = 0.7) 
diat_plot

```

## Clean data and Estimate Principal Curves (PrC) of the F9 pollen core
```{r pollen data F9, echo=FALSE, warning=FALSE}
#Read in F9 pollen record
F_pollen <- read.csv("data/Fuquene_pollen_counts.csv") %>%
  #rename(depth=ï..depth) %>% #rename extraneous column name  
  select(-c(contains("Lycopo") | contains("Tot.LocTot.100")))  

names(F_pollen)

#this is to transform to tidy format, calculate % and subset more common species
pollen_long <- F_pollen %>% 
  gather(key = taxa, value = count, -depth, -upper_age, -lower_age) %>%
  #mutate(taxa = plyr::mapvalues(taxa, from = changes[,1], to = changes$new_2)) %>%
  group_by(depth, taxa, upper_age, lower_age) %>%
  summarise(count = sum(count)) %>%
  filter(!count == "0" ) %>% #this is to remove empty samples (rows)
  filter(!upper_age==0.0) %>% #this is to drop extraneous ages
  ungroup() %>%
  group_by(depth) %>%
  mutate(relative_abundance_percent = count / sum(count) * 100) %>%
  ungroup()

# filter more abundant taxa; format need to be on long-wide format-->no spreaded 
core_common_taxa <- pollen_long %>%
  group_by(taxa) %>%
  summarise(max_rel_abund = max(relative_abundance_percent)) %>%
  filter(max_rel_abund >= 5) %>%
  arrange(max_rel_abund) %>%
  pull(taxa)

# select from initial table
core_counts_common_F7 <- pollen_long %>%
  filter(taxa %in% core_common_taxa) %>%
  mutate(taxa = factor(taxa, levels = core_common_taxa)) %>%
  arrange(taxa)

#make it wide
core_counts_wide_F7 <- core_counts_common_F7 %>%
  select(depth, upper_age, lower_age, taxa, relative_abundance_percent) %>%
  spread(key = taxa, value = relative_abundance_percent) %>%
  arrange(upper_age)

# check age core ranges
range(core_counts_wide_F7$upper_age)

```

### Check temporal resolution of pollen F7 record
```{r}
median(diff(core_counts_wide_F7$upper_age))
plot.ts(diff(core_counts_wide_F7$upper_age))
```

### Plot PrC trends
```{r}
# remove agedepth columns for Hellinger transformation and bind back
agedepth <- core_counts_wide[,names(core_counts_wide) %in% c("depth", "upper_age", "lower_age")]
pollen <- core_counts_wide[,!(names(core_counts_wide) %in% c("upper_age", "lower_age", "depth"))]
pollen[is.na(pollen)] <- 0 #Replace NA (if any) by 0

# Transform data to Hellinger form
pollen_hell <- decostand(pollen, method="hellinger")

# Run Principal Curves and extract scores for then combining the resulting dataframe with agedepth
pollen.prc <- prcurve(pollen_hell, method = "ca", trace = TRUE, vary = TRUE, penalty = 1.4)
scrs_prc <- scores(pollen.prc, display = "curve")

PollenPrC <- cbind(agedepth, scrs_prc)

# Plot Pcurves with depth and ages
PollenPlot <- ggplot(PollenPrC, aes(x = upper_age, y = PrC)) +
  geom_line() + geom_point() +
  labs(y = "Pollen PrC", x = "Age (cal yr BP)", title = "") +
  ggtitle("Agropastoralism PrC")
PollenPlot
```


## Clean data and Estimate Principal Curves (PrC) of the M1 pollen core
```{r pollen data M1, echo=FALSE, warning=FALSE}
M1_pollen_all <- read.csv("data/Fuquene_pollen_M1_counts.csv") %>%
    select(-c(contains("Lycopo") | contains("Carbones") | contains("concentration") | contains("Suma.total") |
                contains("Pollen.spores")))  

# here drop out aquatic pollen and spores
names(M1_pollen_all)
drop <- c(85:105,106,107,108,109,111, 112, 114,115,117,119) #indexes where spores and aquatic pollen are

M1_pollen <- M1_pollen_all %>% 
  select(!all_of(drop)) %>% select(-c(contains("Huevo")))
names(M1_pollen)

# check ages ranges
range(M1_pollen$upper_age)

#this is to transform to tidy format, calculate % and subset more common species
pollen_long <- M1_pollen %>% 
  gather(key = taxa, value = count, -depth, -upper_age, -lower_age) %>%
  #mutate(taxa = plyr::mapvalues(taxa, from = changes[,1], to = changes$new_2)) %>%
  group_by(depth, taxa, upper_age, lower_age) %>%
  summarise(count = sum(count)) %>%
  filter(!count == "0" ) %>% #this is to remove empty samples (rows)
  filter(!upper_age==0.0) %>% #this is to drop extraneous ages
  ungroup() %>%
  group_by(depth) %>%
  mutate(relative_abundance_percent = count / sum(count) * 100) %>%
  ungroup()

# filter more abundant taxa; format need to be on long-wide format-->no spreaded 
core_common_taxa <- pollen_long %>%
  group_by(taxa) %>%
  summarise(max_rel_abund = max(relative_abundance_percent)) %>%
  filter(max_rel_abund >= 3) %>%
  arrange(max_rel_abund) %>%
  pull(taxa)

# select from initial table
core_counts_common_M1 <- pollen_long %>%
  filter(taxa %in% core_common_taxa) %>%
  mutate(taxa = factor(taxa, levels = core_common_taxa)) %>%
  arrange(taxa)

#make it wide
core_counts_wide_M1 <- core_counts_common %>%
  select(depth, upper_age, lower_age, taxa, relative_abundance_percent) %>%
  spread(key = taxa, value = relative_abundance_percent) 

```

## Join M1 and F7 records by common taxa names (columns)
```{r join pollen records, echo=FALSE, message=FALSE, warning=FALSE}
#combine cores (join's analogue package function)
df <- analogue::join(core_counts_wide_M1, core_counts_wide_F7, verbose = TRUE)

#check NA in the list and name the list
listnans <- lapply(df, function(x) sum(is.na(x)))
names(cores) <- c("M1", "F7")

# extract merged cores and sort by age
merged <- plyr::ldply(df, data.frame)
merged <- merged[,-1] #remove .id variable

pollen_M1_F7 <- merged[order(merged$upper_age),] #arrange in increasing age

```

## Stratigraphic plot pollen cores M1 and F7
```{r}
# Gather dataset
pollen_long_M1_F7 <- pollen_M1_F7 %>% 
  gather(key = taxa, value = relative_abundance_percent, -depth, -upper_age, -lower_age) 

# Plot  
pollen_plot <- ggplot(pollen_long_M1_F7, aes(x = relative_abundance_percent, y = upper_age)) +
  geom_col_segsh() +
  scale_y_reverse() +
  facet_abundanceh(vars(taxa)) +
  labs(x = "Relative abundance (%)", y = "Cal yr BP")
  # add CONISS zones
  #geom_hline(yintercept = zones, col = "blue", lty = 1, alpha = 0.7) 
pollen_plot
```





