---
title: "WORKFLOW for data analyses of Fuquene record"
author: 'Author: Xavier Benito (xavier.benito.granell@gmail.com)'
date: 'Date: 11/05/2022'
output:
  html_notebook: default
  pdf_document: default
---

## Introduction
This workflow is written as a R Markdown document to show initial exploratory analyses of paleoclimatic and paleoenvironmental evolution of Lake Fuquene (Colombia, 2540m elevation, 5°27′N, 73°46′W). This document shows the most relevant code. See R files within the folder "scripts" for a more specific code. 

### Study area and datasets description
+ **Lake Fuquene**: Pleistocene in origin, morrained-dammed tropical lake located in the Eastern Cordiellera of Colombian Andes. Water depth=2-6 m

### Three sediment cores, composite of 250 kyr diatom, geochemical and pollen record
+ **F9**: 60-m long, sampled for diatom, pollen and geochemistry
+ **F7**: sampled for diatoms
+ **M1**: sampled for diatoms and pollen


## Main numerical steps:
+ Clean data and estimate Principal Curves for each assemblage (pollen and diatoms)
+ Plot stratigraphical floristic changes over time to identify periods of change
+ Interpolate PrC from coarser to lower resolution 
+ GAM covariates model
+ HGAM-derivatives: assemblage-wide rate of change


### Load packages 
```{r packages, echo=TRUE, message=FALSE, warning=FALSE}
library(analogue)
library(tidyverse)
library(vegan)
library(analogue)
library(mgcv)
library(ggplot2)
library(gratia)
```

## Step 1: Clean data and estimate Principal Curves (PrC) of the diatom cores (M1, F7 and F9)
PrC is a nonlinear ordination technique that extracts a single gradient of variation from multivariate data.

```{r principal curves diatoms, echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
F_diatoms <- read.csv("data/Fuquene_diatoms_counts.csv") %>%
  rename(core=ï..core) #rename extraneous column name 

agedepth <- F_diatoms[, names(F_diatoms) %in% c("depth", "upper_age", "lower_age", "core")] 
diat <- F_diatoms[, !names(F_diatoms) %in% c("depth", "upper_age", "lower_age", "core")]
diat[is.na(diat)] <- 0

diatoms_save <- cbind(agedepth, diat)

#this is to transform to tidy format, calculate % and subset more common species
new <- diatoms_save %>% 
  gather(key = taxa, value = count, -depth, -upper_age, -lower_age, -core) %>%
  #mutate(taxa = plyr::mapvalues(taxa, from = changes[,1], to = changes$new_2)) %>%
  group_by(depth, taxa, upper_age, lower_age) %>%
  summarise(count = sum(count)) %>%
  filter(!count == "0" ) %>% #this is to remove empty samples (rows)
  filter(!upper_age==0.0) %>% #this is to drop extraneous ages
  ungroup() %>%
  group_by(depth) %>%
  mutate(relative_abundance_percent = count / sum(count) * 100) %>%
  mutate(elapsedTime = abs(upper_age - lower_age)) %>%
  mutate(negAge=-upper_age) %>%
  ungroup() 


# filter more abundant taxa; format need to be on long-wide format-->no spreaded 
core_common_taxa <- new %>%
  group_by(taxa) %>%
  summarise(max_rel_abund = max(relative_abundance_percent)) %>%
  filter(max_rel_abund >= 5) %>%
  arrange(max_rel_abund) %>%
  pull(taxa)

# select from initial table
core_counts_common <- new %>%
  filter(taxa %in% core_common_taxa) %>%
  mutate(taxa = factor(taxa, levels = core_common_taxa)) %>%
  arrange(taxa)

#make it wide
core_counts_wide_diatoms <- core_counts_common %>%
  select(depth, upper_age, lower_age, taxa, relative_abundance_percent) %>%
  spread(key = taxa, value = relative_abundance_percent) %>%
  arrange(upper_age) #sort by increasing time
```

### Check temporal resolution of diatom record
```{r}
median(diff(core_counts_wide_diatoms$upper_age))
plot.ts(diff(core_counts_wide_diatoms$upper_age))
```

### Estime PrC and plot trends
```{r PrC diatoms, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
## extract agedepth variables
agedepth <- core_counts_wide_diatoms[,names(core_counts_wide_diatoms) %in% c("depth", "upper_age", "lower_age")]
diatoms <- core_counts_wide_diatoms[,!(names(core_counts_wide_diatoms) %in% c("upper_age", "lower_age", "depth"))]

# Transform data to Hellinger form
diatoms[is.na(diatoms)] <- 0 #Replace NA (if any) by 0
diat_hell <- decostand(diatoms, method="hellinger")

# Run Principal Curves
diat_prc <- prcurve(diat_hell, method = "ca", trace = TRUE, vary = TRUE, penalty = 1.4)

## Extract position on the curve
scrs_prc <- scores(diat_prc, display = "curve")

# Combine dataframe with ages and depths
diatomsPrC <- cbind(agedepth, scrs_prc)

# Plot Pcurves with depth and ages
diat_plt_prc <- ggplot(diatomsPrC, aes(x = upper_age, y = PrC)) +
  geom_line() + geom_point() +
  labs(y = "PrC", x = "Age (cal yr BP)", title = "") +
  ggtitle("Diatoms PrC") +
  theme_bw()
diat_plt_prc

```

## Stratigraphic diatom plot

### First do CONISS classification to see statistically significant stratigraphic zones
```{r CONISS diatom, echo=FALSE, message=FALSE, warning=FALSE}
library(rioja)

#do coniss to add statistically significant stratigraphic zones
core_counts_wide[is.na(core_counts_wide)] <- 0
      
diatHel <- decostand(core_counts_wide[,4:ncol(core_counts_wide)], method="hellinger")
diss <- vegdist(diatHel, method="bray")
clust <- chclust(diss, method="coniss")
bstick(clust)

#doesn't work
zones <- cutree(clust, k=8)
locate <- cumsum(rle(zones)$lengths)+1
zones <- core_counts_wide_diatoms[locate, ][,2]
zones <- zones$upper_age

```

```{r stratiplot, results='markup'}
## Using tidypaleo R package (https://fishandwhistle.net/post/2018/stratigraphic-diagrams-with-tidypaleo-ggplot2/)
library(tidypaleo) #remotes::install_github("paleolimbot/tidypaleo")
library(patchwork)

theme_set(theme_bw(9))

# Plot  
diat_plot <- ggplot(core_counts_common, aes(x = relative_abundance_percent, y = upper_age)) +
  geom_col_segsh() +
  scale_y_reverse() +
  facet_abundanceh(vars(taxa)) +
  labs(x = "Relative abundance (%)", y = "Cal yr BP") +
  # add CONISS zones
  geom_hline(yintercept = zones, col = "blue", lty = 1, alpha = 0.7) 
diat_plot

```

## Clean data and Estimate Principal Curves (PrC) of the F9 pollen core
```{r pollen data F9, echo=FALSE, warning=FALSE}
#Read in F9 pollen record
F_pollen <- read.csv("data/Fuquene_pollen_counts.csv") %>%
  #rename(depth=ï..depth) %>% #rename extraneous column name  
  select(-c(contains("Lycopo") | contains("Tot.LocTot.100"))) %>%
  select(-c(66,67,68,70)) %>% 
  rename(depth=ï..depth)

names(F_pollen)

#this is to transform to tidy format, calculate % and subset more common species
pollen_long <- F_pollen %>% 
  gather(key = taxa, value = count, -depth, -upper_age, -lower_age) %>%
  #mutate(taxa = plyr::mapvalues(taxa, from = changes[,1], to = changes$new_2)) %>%
  group_by(depth, taxa, upper_age, lower_age) %>%
  summarise(count = sum(count)) %>%
  filter(!count == "0" ) %>% #this is to remove empty samples (rows)
  filter(!upper_age==0.0) %>% #this is to drop extraneous ages
  ungroup() %>%
  group_by(depth) %>%
  mutate(relative_abundance_percent = count / sum(count) * 100) %>%
  ungroup()

# filter more abundant taxa; format need to be on long-wide format-->no spreaded 
core_common_taxa <- pollen_long %>%
  group_by(taxa) %>%
  summarise(max_rel_abund = max(relative_abundance_percent)) %>%
  filter(max_rel_abund >= 5) %>%
  arrange(max_rel_abund) %>%
  pull(taxa)

# select from initial table
core_counts_common_F9 <- pollen_long %>%
  filter(taxa %in% core_common_taxa) %>%
  mutate(taxa = factor(taxa, levels = core_common_taxa)) %>%
  arrange(taxa)

#make it wide
core_counts_wide_F9 <- core_counts_common_F9 %>%
  select(depth, upper_age, lower_age, taxa, relative_abundance_percent) %>%
  spread(key = taxa, value = relative_abundance_percent) %>%
  arrange(upper_age)

# check age core ranges
range(core_counts_wide_F9$upper_age)

```

### Check temporal resolution of pollen F9 record
```{r}
mean(diff(core_counts_wide_F9$upper_age))
plot.ts(diff(core_counts_wide_F9$upper_age))
```

### Plot PrC trends
```{r}
# remove agedepth columns for Hellinger transformation and bind back
agedepth <- core_counts_wide_F9[,names(core_counts_wide_F9) %in% c("depth", "upper_age", "lower_age")]
pollen <- core_counts_wide_F9[,!(names(core_counts_wide_F9) %in% c("upper_age", "lower_age", "depth"))]
pollen[is.na(pollen)] <- 0 #Replace NA (if any) by 0

# Transform data to Hellinger form
pollen_hell <- decostand(pollen, method="hellinger")

# Run Principal Curves and extract scores for then combining the resulting dataframe with agedepth
pollen.prc <- prcurve(pollen_hell, method = "ca", trace = TRUE, vary = TRUE, penalty = 1.4)
scrs_prc <- scores(pollen.prc, display = "curve")

PollenPrC <- cbind(agedepth, scrs_prc)

# Plot Pcurves with depth and ages
PollenPlot <- ggplot(PollenPrC, aes(x = upper_age, y = PrC)) +
  geom_line() + geom_point() +
  labs(y = "Pollen PrC", x = "Age (cal yr BP)", title = "") +
  ggtitle("Pollen PrC")
PollenPlot
```


## Clean data and Estimate Principal Curves (PrC) of the M1 pollen core
```{r pollen data M1, echo=FALSE, warning=FALSE}
M1_pollen_all <- read.csv("data/Fuquene_pollen_M1_counts.csv") %>%
    select(-c(contains("Lycopo") | contains("Carbones") | contains("concentration") | contains("Suma.total") |
                contains("Pollen.spores"))) %>%
    rename(depth=ï..depth) #rename extraneous column name 


# here drop out algae and spores
names(M1_pollen_all)
drop <- c(85:119) #indexes where spores and algae are

M1_pollen <- M1_pollen_all %>% 
  select(!all_of(drop)) %>% select(-c(contains("Huevo")))
names(M1_pollen)

# check ages ranges
range(M1_pollen$upper_age)

#this is to transform to tidy format, calculate % and subset more common species
pollen_long <- M1_pollen %>% 
  gather(key = taxa, value = count, -depth, -upper_age, -lower_age) %>%
  #mutate(taxa = plyr::mapvalues(taxa, from = changes[,1], to = changes$new_2)) %>%
  group_by(depth, taxa, upper_age, lower_age) %>%
  summarise(count = sum(count)) %>%
  filter(!count == "0" ) %>% #this is to remove empty samples (rows)
  filter(!upper_age==0.0) %>% #this is to drop extraneous ages
  ungroup() %>%
  group_by(depth) %>%
  mutate(relative_abundance_percent = count / sum(count) * 100) %>%
  ungroup()

# filter more abundant taxa; format need to be on long-wide format-->no spreaded 
core_common_taxa <- pollen_long %>%
  group_by(taxa) %>%
  summarise(max_rel_abund = max(relative_abundance_percent)) %>%
  filter(max_rel_abund >= 5) %>%
  arrange(max_rel_abund) %>%
  pull(taxa)

# select from initial table
core_counts_common_M1 <- pollen_long %>%
  filter(taxa %in% core_common_taxa) %>%
  mutate(taxa = factor(taxa, levels = core_common_taxa)) %>%
  arrange(taxa)

#make it wide
core_counts_wide_M1 <- core_counts_common_M1 %>%
  select(depth, upper_age, lower_age, taxa, relative_abundance_percent) %>%
  spread(key = taxa, value = relative_abundance_percent) 

```

## Join M1 and F7 records by common taxa names (columns)

```{r join pollen records, echo=FALSE, message=FALSE, warning=FALSE}
#combine cores (join's analogue package function)
df <- analogue::join(core_counts_wide_M1, core_counts_wide_F9, verbose = TRUE)

#check NA in the list and name the list
listnans <- lapply(df, function(x) sum(is.na(x)))
names(df) <- c("M1", "F7")

# extract merged cores and sort by age
merged <- plyr::ldply(df, data.frame)
# merged <- merged[,-1] #remove .id variable

pollen_M1_F9 <- merged[order(merged$upper_age),] %>% #arrange in increasing age
                  rename(core=.id)

```

## Stratigraphic plot pollen cores M1 and F7
```{r}
# Gather dataset
pollen_long_M1_F9 <- pollen_M1_F9 %>% 
  gather(key = taxa, value = relative_abundance_percent, -depth, -upper_age, -lower_age, -core) 

# Plot  
pollen_plot <- ggplot(pollen_long_M1_F9, aes(x = relative_abundance_percent, y = upper_age)) +
  geom_col_segsh() +
  scale_y_reverse() +
  facet_abundanceh(vars(taxa)) +
  labs(x = "Relative abundance (%)", y = "Cal yr BP")
  # add CONISS zones
  #geom_hline(yintercept = zones, col = "blue", lty = 1, alpha = 0.7) 
pollen_plot
```

## Geochemical and gran size datasets

```{r}
#Read in geochemistry 
F9_geochem <- read.csv("data/Fuquene_F9_geochem.csv") %>%
  rename(depth=ï..depth) %>%
  rename(TIC_percent=X..TIC) %>%
  rename(N_percent=wt..N) %>%
  rename(C_percent=wt..C) %>%
  rename(TOC_percent=wt..TOC) %>%
  rename(TC_TN=TC.TN) %>%
  rename(d13C=d13C..permil..vs.VPDB.)

names(F9_geochem)
 
#check median temporal resolution
median(diff(F9_geochem$upper_age))
plot.ts(diff(F9_geochem$upper_age))

#Read in grain size 
F9_grainsize <- read.csv("data/Fuquene_F9_grainsize.csv") %>%
  rename(depth=depth..cm.) %>%
  rename(upper_age=ï..upper_age)

#check median temporal resolution
median(diff(F9_grainsize$upper_age))
plot.ts(diff(F9_grainsize$upper_age))

```

```{r}
### Interpolation to the coarser dataset: geochemistry

LOI375_i <- as.data.frame(approx(F9_grainsize$upper_age, F9_grainsize$LOI.375, F9_geochem$upper_age)$y)
LOI375_i$age <- F9_geochem$upper_age
colnames(LOI375_i) <- c("LOI375", "age")

LOI375_i$sand <- approx(F9_grainsize$upper_age, F9_grainsize$EM1, F9_geochem$upper_age)$y
LOI375_i$coarsesilt <- approx(F9_grainsize$upper_age, F9_grainsize$EM2, F9_geochem$upper_age)$y
LOI375_i$finesilt <- approx(F9_grainsize$upper_age, F9_grainsize$EM3, F9_geochem$upper_age)$y
LOI375_i$clay <- approx(F9_grainsize$upper_age, F9_grainsize$EM4, F9_geochem$upper_age)$y

F9_geochem_grainsize_i <- LOI375_i

head(F9_geochem_grainsize_i)
diff(F9_geochem_grainsize_i$age)

#plot(diatom_to_pollen_prc_i[,1], diatom_to_pollen_prc_i[,3])
```


## Hierarchical Generalized Additive Models (HGAM) -- **Synchronous model**
In this step we model, first, diatom and pollen assemblages using HGAM and then calculate average rate of change at assemblage-level to assess the timing in the start of rapid change in the time-series, and test comparable rates of change between aquatic and terrestrial ecosystems.

By obtaining derivative-based values of the diatom, terrestrial pollen, and agropastoralism time series we obtain regular time intervals at XX-year resolution (median between diatom and pollen records) and hence comparable--same time steps-- for fitting the synchronous and asynchronous models. 

We fit model S HGAM (Pedersen et al., 2019): group-level smoothness between groups (spp) without a global smooth while allowing different willingness to not assume any shape among species for each assemblage.

### Diatom assemblages
This chunk is reproducible for 

```{r}

#Prepare the data
agedepth <- F_diatoms[, names(F_diatoms) %in% c("depth", "upper_age", "lower_age", "core")] 
diat <- F_diatoms[, !names(F_diatoms) %in% c("depth", "upper_age", "lower_age", "core")]
diat[is.na(diat)] <- 0
agedepth$elapsed <- round(abs(agedepth$upper_age - agedepth$lower_age),0)

#Select most common species 
criteria <- 0.2 #% of the total samples

n.occur <- apply(diat>0, 2, sum)
diat_red <- diat[, n.occur > (dim(diat)[1])*criteria] #
diatoms <- cbind(agedepth, diat_red)

#this is to transform to tidy format, calculate % and subset more common species
diat_data <- diatoms %>% 
  gather(key = taxa, value = count, -depth, -upper_age, -lower_age, -core) %>%
  #mutate(taxa = plyr::mapvalues(taxa, from = changes[,1], to = changes$new_2)) %>%
  group_by(depth) %>%
  mutate(total_sample = sum(count)) %>% 
  filter(!total_sample == "0") %>% #this is to remove empty samples
  mutate(log_total_counts = log10(total_sample+1)) %>%
  mutate(relative_abundance_percent = count / sum(count) * 100) %>%
  mutate(negAge = -upper_age) %>%
  #mutate(AgeCE = upper_age*(-1)+1950) %>%
  #filter(AgeCE >= "0") %>%
  mutate(elapsedTime = round(abs(upper_age - lower_age),0)) %>%
  ungroup() %>%
  filter(!is.na(elapsedTime)) %>%
  mutate(spp = factor(taxa)) 

levels(diat_data$spp)

#model S HGAM : similar smootheness between groups (spp) without global smooth 
set.seed(10) #set a seed so this is repeatable
diatom_gam_S <- gam(count ~ s(negAge, spp, k=20, bs="fs") + offset(log_total_counts),
                    weights = elapsedTime/mean(elapsedTime),
                    data=diat_data, family = nb, #places notes at the deciles of sample ages
                    method = "REML")

summary(diatom_gam_S)
gam.check(diatom_gam_S)
draw(diatom_gam_S)

#model I HGAM: different smootheness for each taxa without global smooth
diatom_gam_I<- gam(count ~ s(negAge, by=spp, k=20, bs="fs") +
                     s(spp, bs="re") + offset(log_total_counts),
                   weights = elapsedTime/mean(elapsedTime),
                   data=diat_data, family = nb, #places notes at the deciles of sample ages
                   method = "REML")

gam.check(diatom_gam_I)
draw(diatom_gam_I)

#Compare different model fits using AIC
AIC_table <- AIC(diatom_gam_S, diatom_gam_I)%>%
  rownames_to_column(var= "Model")%>%
  mutate(data_source = rep(c("diatom_data")))%>%
  group_by(data_source)%>%
  mutate(deltaAIC = AIC - min(AIC))%>%
  ungroup()%>%
  dplyr::select(-data_source)%>%
  mutate_at(.vars = vars(df,AIC, deltaAIC), 
            .funs = funs(round,.args = list(digits=0)))

#Create synthetic data to predict over a range of ages
diat_plot_data <- with(diat_data, as_tibble(expand.grid(negAge = seq(min(diat_data$negAge), max(diat_data$negAge)),
                                                        spp = factor(levels(diat_data$spp)),
                                                        log_total_counts = mean(log_total_counts))))

diat_modS_fit <- predict(diatom_gam_S, 
                         newdata = diat_plot_data,
                         se.fit = TRUE)

diat_modI_fit <- predict(diatom_gam_I,
                         newdata = diat_plot_data,
                         se.fit = TRUE)

#non-shared trends
diat_plot_data$modS_fit <- as.numeric(diat_modS_fit$fit)
diat_plot_data$modI_fit <- as.numeric(diat_modI_fit$fit)

# comparing non-shared trends
diat_plot_data <- gather(diat_plot_data, key=model, value=fit, modS_fit, modI_fit)

diat_plot_data <- mutate(diat_plot_data, se= c(as.numeric(diat_modS_fit$se.fit),
                                               as.numeric(diat_modI_fit$se.fit)),
                         upper = exp(fit + (2 * se)),
                         lower = exp(fit - (2 * se)),
                         fit   = exp(fit))

# For only one model
# diat_plot_data <- gather(diat_plot_data, key=model, value=fit, modI_fit)
# diat_plot_data <- mutate(diat_plot_data, se= c(as.numeric(diat_modI_fit$se.fit)),
#                          upper = exp(fit + (2 * se)),
#                          lower = exp(fit - (2 * se)),
#                          fit   = exp(fit))

#Plot the model output for non-shared trends, with means plus standard deviations for each model.
diat_plot_model_labels <- paste("Model", c("S", "I"))
diat_plot_model_labels <- factor(diat_plot_model_labels, levels = diat_plot_model_labels)

#non-shared trends
theme_set(theme_bw())
theme_update(panel.grid = element_blank())

diat_plot <- ggplot(diat_plot_data) +
  facet_wrap(~spp, nrow = 4,scales = "free_y")+
  geom_ribbon(aes(x=negAge,
                  ymin = lower,
                  ymax = upper,
                  fill = model),
              alpha=0.2)+
  geom_point(data= diat_data, aes(x = negAge, y = count), size=0.06) +
  geom_line(aes(x = negAge, y = fit, color = model))+
  labs(y = "Absolute counts", x = "Age (cal yr BP)") +
  scale_fill_brewer(name = "", palette = "Dark2",
                    labels = diat_plot_model_labels) +
  scale_colour_brewer(name = "",
                      palette = "Dark2", labels = diat_plot_model_labels)+
  theme(legend.position = "top",
        strip.text = element_text(size=10))

diat_plot

## save model results for later use
write.csv(diat_plot_data, "outputs/diatoms-HGAMs-fitted-values.csv", row.names = FALSE)

```


##Derivatives and posterior distribution simulation

```{r}
# Eric's code
#Function for calculating first derivatives of time series given a before,
#after, and delta step size
calc_1st_deriv = function(fit_before, fit_after,delta) {
  (fit_after-fit_before)/(2*delta)
}

set.seed(10) #set a seed so this is repeatable
n_sims = 250

n_length = 240

years <- seq(min(diat_plot_data$negAge),
             max(diat_plot_data$negAge),
             length.out = n_length)

diff(years)
#model <- diatom_gam_S

model <- diatom_gam_I

#pred <- diat_modS_fit
pred <- diat_modI_fit

# Generate multivariate normal simulations of model coefficients
random_coefs <- t(rmvn(n_sims, mu = coef(model),V = vcov(model)))

confint_sims <- crossing(spp=unique(diat_plot_data$spp),
                         negAge = seq(min(diat_plot_data$negAge),
                                      max(diat_plot_data$negAge),
                                      length.out = n_length),
                         log_total_counts=0)

map_pred_sims <- predict(model,
                         confint_sims,
                         type = "lpmatrix") %*% random_coefs %>%
  as_data_frame() %>%
  bind_cols(confint_sims)%>%
  gather(key = simulation, value = pred, -negAge, -log_total_counts,-spp)


#specifying the step size for numerical derivative calculations
delta = 0.01

#calculating the predicted value for the current year plus delta
step_ahead_fits = confint_sims %>%
  mutate(negAge = negAge+delta)%>%
  predict(model, 
          ., type = "lpmatrix") %*% random_coefs 


#calculating the predicted value for the current year minus delta
step_behind_fits = confint_sims %>%
  mutate(negAge = negAge-delta)%>%
  predict(model,
          ., type = "lpmatrix") %*% random_coefs 


#using the predicted values for year plus and minus delta to calculate
#derivatives for each species for each simulation
derivs <- calc_1st_deriv(step_behind_fits,step_ahead_fits,delta = delta)%>%
  as_data_frame()%>%
  bind_cols(confint_sims)%>%
  gather(key = simulation,value = deriv, -spp,-negAge, -log_total_counts)

#Creating summaries of derivatives for each simulation for each year
deriv_summaries <- derivs %>%
  group_by(negAge,simulation)%>%
  summarize(deriv_mean = mean(deriv),
            deriv_sd = sd(deriv))%>%
  group_by(negAge)%>% #turning derivative summaries into 95% confidence intervals
  select(-simulation)%>%
  summarize_all(.funs = list(lower = ~quantile(.,probs = 0.025),
                             upper = ~quantile(.,probs = 0.975),
                             med   = ~quantile(.,probs = 0.5)))

#Plotting mean rate of change plus the 95% CI
mean_plot <- deriv_summaries %>%
  ggplot(aes(x = negAge, 
             y = deriv_mean_med, 
             ymin = deriv_mean_lower,
             ymax = deriv_mean_upper))+
  geom_ribbon(fill="grey")+
  geom_line()+
  geom_hline(yintercept = 0, linetype=2) +
  scale_y_continuous("")+
  xlab("Cal years BP") +
  theme_bw()
mean_plot 

#Plotting standard deviation of rate of change plus the 95% CI
sd_plot <- deriv_summaries %>%
  ggplot(aes(x = negAge, 
             y = deriv_sd_med, 
             ymin=deriv_sd_lower,
             ymax=deriv_sd_upper))+
  geom_ribbon(fill="grey")+
  geom_line()+
  geom_hline(yintercept = 0, linetype=2) +
  scale_y_continuous("")+
  xlab("Cal years BP") +
  theme_bw()
sd_plot


## save derivative summaries for later use
write.csv(deriv_summaries, "outputs/diatom-derivatives.csv", row.names = FALSE)
```


