---
title: "WORKFLOW for data analyses of Fuquene record"
author: 'Author: Xavier Benito (xavier.benito.granell@gmail.com)'
date: 'Date: 11/05/2022'
output:
  html_notebook: default
  pdf_document: default
---

## Introduction
This workflow is written as a R Markdown document to show initial exploratory analyses of paleoclimatic and paleoenvironmental evolution of Lake Fuquene (Colombia, 2540m elevation, 5°27′N, 73°46′W). This document shows the most relevant code. See R files within the folder "scripts" for a more specific code. 

## Study area and datasets description
+ **Lake Fuquene**: Pleistocene in origin, morrained-dammed tropical lake located in the Eastern Cordiellera of Colombian Andes. Water depth=2-6 m

Three sediment cores, composite of 250 kyr diatom, geochemical and pollen record:
  + **F9**: 60-m long, sampled for diatom, pollen and geochemistry
  + **F7**: sampled for diatoms
  + **M1**: sampled for diatoms and pollen


## Load packages 
```{r packages, include=FALSE}
library(analogue)
library(tidyverse)
library(vegan)
library(analogue)
library(mgcv)
library(ggplot2)
library(gratia)
```


## Step 1: Estimate Principal Curves (PrC) of the diatom data
PrC is a nonlinear ordination technique that extracts a single gradient of variation from multivariate data.

# Clean data
```{r principal curves diatoms, echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}

F_diatoms <- read.csv("data/Fuquene_diatoms_counts.csv") %>%
  rename(core=ï..core) #rename extraneous column name 

agedepth <- F_diatoms[, names(F_diatoms) %in% c("depth", "upper_age", "lower_age", "core")] 
diat <- F_diatoms[, !names(F_diatoms) %in% c("depth", "upper_age", "lower_age", "core")]
diat[is.na(diat)] <- 0

diatoms_save <- cbind(agedepth, diat)

#this is to transform to tidy format, calculate % and subset more common species
new <- diatoms_save %>% 
  gather(key = taxa, value = count, -depth, -upper_age, -lower_age, -core) %>%
  #mutate(taxa = plyr::mapvalues(taxa, from = changes[,1], to = changes$new_2)) %>%
  group_by(depth, taxa, upper_age, lower_age) %>%
  summarise(count = sum(count)) %>%
  filter(!count == "0" ) %>% #this is to remove empty samples (rows)
  filter(!upper_age==0.0) %>% #this is to drop extraneous ages
  ungroup() %>%
  group_by(depth) %>%
  mutate(relative_abundance_percent = count / sum(count) * 100) %>%
  ungroup()

# filter more abundant taxa; format need to be on long-wide format-->no spreaded 
core_common_taxa <- new %>%
  group_by(taxa) %>%
  summarise(max_rel_abund = max(relative_abundance_percent)) %>%
  filter(max_rel_abund >= 5) %>%
  arrange(max_rel_abund) %>%
  pull(taxa)

# select from initial table
core_counts_common <- new %>%
  filter(taxa %in% core_common_taxa) %>%
  mutate(taxa = factor(taxa, levels = core_common_taxa)) %>%
  arrange(taxa)

#make it wide
core_counts_wide <- core_counts_common %>%
  select(depth, upper_age, lower_age, taxa, relative_abundance_percent) %>%
  spread(key = taxa, value = relative_abundance_percent) 

```

# Estime PrC and plot trends

```{r PrC, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
## extract agedepth variables
agedepth <- core_counts_wide[,names(core_counts_wide) %in% c("depth", "upper_age", "lower_age")]
diatoms <- core_counts_wide[,!(names(core_counts_wide) %in% c("upper_age", "lower_age", "depth"))]

# Transform data to Hellinger form
diatoms[is.na(diatoms)] <- 0 #Replace NA (if any) by 0
diat_hell <- decostand(diatoms, method="hellinger")

# Run Principal Curves
diat_prc <- prcurve(diat_hell, method = "ca", trace = TRUE, vary = TRUE, penalty = 1.4)

## Extract position on the curve
scrs_prc <- scores(diat_prc, display = "curve")

# Combine dataframe with ages and depths
diatomsPrC <- cbind(agedepth, scrs_prc)

# Plot Pcurves with depth and ages
diat_plt_prc <- ggplot(diatomsPrC, aes(x = upper_age, y = PrC)) +
  geom_line() + geom_point() +
  labs(y = "PrC", x = "Age (cal yr BP)", title = "") +
  ggtitle("Diatoms PrC") +
  theme_bw()
diat_plt_prc

```

## Step 2: Estimate Principal Curves (PrC) of the pollen data
```{r pollen data, echo=FALSE, warning=FALSE}
F_pollen <- read.csv("data/Fuquene_pollen_counts.csv") %>%
  rename(depth=ï..depth) %>% #rename extraneous column name  
  select(-c(contains("Lycopo") | contains("Tot.LocTot.100")))  

names(F_pollen)

#this is to transform to tidy format, calculate % and subset more common species
pollen_long <- F_pollen %>% 
  gather(key = taxa, value = count, -depth, -upper_age, -lower_age) %>%
  #mutate(taxa = plyr::mapvalues(taxa, from = changes[,1], to = changes$new_2)) %>%
  group_by(depth, taxa, upper_age, lower_age) %>%
  summarise(count = sum(count)) %>%
  filter(!count == "0" ) %>% #this is to remove empty samples (rows)
  filter(!upper_age==0.0) %>% #this is to drop extraneous ages
  ungroup() %>%
  group_by(depth) %>%
  mutate(relative_abundance_percent = count / sum(count) * 100) %>%
  ungroup()

# filter more abundant taxa; format need to be on long-wide format-->no spreaded 
core_common_taxa <- pollen_long %>%
  group_by(taxa) %>%
  summarise(max_rel_abund = max(relative_abundance_percent)) %>%
  filter(max_rel_abund >= 5) %>%
  arrange(max_rel_abund) %>%
  pull(taxa)

# select from initial table
core_counts_common <- pollen_long %>%
  filter(taxa %in% core_common_taxa) %>%
  mutate(taxa = factor(taxa, levels = core_common_taxa)) %>%
  arrange(taxa)

#make it wide
core_counts_wide <- core_counts_common %>%
  select(depth, upper_age, lower_age, taxa, relative_abundance_percent) %>%
  spread(key = taxa, value = relative_abundance_percent) 

agedepth <- core_counts_wide[,names(core_counts_wide) %in% c("depth", "upper_age", "lower_age")]
pollen <- core_counts_wide[,!(names(core_counts_wide) %in% c("upper_age", "lower_age", "depth"))]
pollen[is.na(pollen)] <- 0 #Replace NA (if any) by 0

# Transform data to Hellinger form
pollen_hell <- decostand(pollen, method="hellinger")

# Run Principal Curves and extract scores for then combining the resulting dataframe with agedepth
pollen.prc <- prcurve(pollen_hell, method = "ca", trace = TRUE, vary = TRUE, penalty = 1.4)
scrs_prc <- scores(pollen.prc, display = "curve")

PollenPrC <- cbind(agedepth, scrs_prc)

# Plot Pcurves with depth and ages
PollenPlot <- ggplot(PollenPrC, aes(x = upper_age, y = PrC)) +
  geom_line() + geom_point() +
  labs(y = "Pollen PrC", x = "Age (cal yr BP)", title = "") +
  ggtitle("Agropastoralism PrC")
PollenPlot

```

